Quantifying the impacts of bias in landcover data on global change analyses

### 2015-08-30 paper/main.tex
Main paper: Half of results done, bullet points on rest through Discussion (but not yet organized).  Updated cropland bias, carbon bias, yield bias, agent-bias, as worked through results. Earlier results reconfirmed.

\vspace{-0.5 cm}
\section{Results}
\subsection{Bias and its correlates}

We created the 1 km reference and landcover maps, removing all croplands marked as communal or smallholder farmland in the reference vector maps (individual fields were not mapped with the same precision) or as plantation forestry (SI), and then aggregated each map to 5, 10, 25, 50, and 100 km resolutions. We then subtracted each landcover map from the reference map at each scale of aggregation (Fig. 1).

\vspace{-0.25 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/bias_mapc.pdf}}
\caption{Differences in percent cropland estimates between the reference map and each of the four landcover maps. Rows indicate the landcover map being assessed (by subtraction from the reference map), while columns refer to resolution of aggregation. White indicates areas with no data where communal farmlands or plantation forests were removed.}\label{afoto}
\end{figure}

The spatial patterns of bias are most pronounced in the MODIS and GlobCover maps, with both substantially underestimating the amount of cropland in most regions of the country, particularly in the central part of the country where the bulk of the country's cropland is found (blue areas in Fig. 1). The mean of these differences, which is the landcover map's bias, are 21\% and 34\% at 1 km for MODIS and GlobCover, respectively (Fig. S1), meaning that each map underestimates cropland by that amount at this resolution. MODIS bias drops to 8\% at 50 km of aggregation, whereas GlobCover bias is still 24\% at 100 km (Fig. S1).

The SA-LC and GeoWiki datasets show much spatial variability and overall bias.

\subsection{Error as a function of cropland density}

\subsection{Potential bias in harvested areas, yield, and production estimates}

\subsection{Potential bias in estimates of carbon stocks}

\subsection{Potential bias in harvested areas, yield, and production estimates}

\subsection{Impacts on evapotranspiration estimates}

@@

@@

\section{Discussion}
\subsection{Blather}
% $\theta$ changes from 0 to 1. (see Figure \ref{afoto}).That means we are considering $\theta$  of the form
%For these solutions we have the following

@@

>>>>

\vspace{-0.5 cm}
\section{Results}
\subsection{Landcover bias}

We created the 1 km reference and landcover maps, removing all croplands marked as communal or smallholder farmland in the reference vector maps (individual fields were not mapped with the same precision), as horticulture, or plantation forestry (SI), and calculated the total cropland extent in the remaining area (1,081,000 km$^2$, or 90\% of South Africa).  The 2011 reference map showed a cropland area of 104,310 km$^2$, which the SA-LC and GeoWiki maps overestimated by 26 and 5.8\%, respectively, and GlobCover and MODIS underestimated by 21 and 26.1\%.

We then aggregated each map to 5, 10, 25, 50, and 100 km resolutions, and subtracted the four landcover maps from the reference map at each scale of aggregation to asses the spatial patterns of bias (Fig. 1).

\vspace{-0.25 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/bias_map.pdf}}
\caption{Differences in percent cropland estimates between the reference map and each of the four landcover maps. Rows indicate the landcover map being assessed (by subtraction from the reference map), while columns refer to resolution of aggregation. White indicates areas with no data where communal farmlands or plantation forests were removed.}\label{afoto}
\end{figure}

The most pronounced extreme biases are in the MODIS and GlobCover maps, with both underestimating cropland extent by 10-75\% in the center of the country (blue areas in Fig. 1, and the dominant production region), and overestimating along the eastern to northern margins (red areas in Fig. 1). The average biases for GlobCover and MODIS are 21\% and 34\% at 1 km, meaning that each map underestimates cropland by that amount at this resolution. MODIS mean bias drops to 8\% at 50 km of aggregation, whereas GlobCover bias is still 24\% at 100 km (SI Appendix, Fig. S1).

The SA-LC map uniformly overestimates cropland throughout the country (Fig. 1), but the magnitude of mean bias is much lower at -8\% at 1 km to -6\% at 100 km  (Fig. S1). The GeoWiki map has a very heterogeneous pattern of bias, having a slight underestimation tendency (5\% mean bias) at 1 km, which changes to a smaller overestimation bias (-2 \%) at 100 km (Fig. S1).

\subsection{The role of landscape characteristics in shaping bias}
To investigate the degree to which landscape features influence landcover map bias, we extract all pixels in agricultural areas ($>$0.5\% cropland) of the 1 km reference map using the the boundaries of 354 magisterial districts (South Africa's finest administrative unit, which average 3,445 km$^2$ in size; SI Appendix, Fig. S3), and averaged the values of these pixels within each district. These provided an estimate of the intensity of cropping within the farmed parts of the landscape (with districts defining the landscape scale), as well as the degree of mixing between cropland and other land covers. We then extracted the cropland map biases for the same locations, and calculated the mean absolute bias within each district. Absolute bias describes the magnitude of bias but not its direction, and is more informative about the likelihood of the map being biased at any given point in the landscape than actual bias, as positive and negative biases can cancel each other out within small areas.

A generalized additive model fit to district-level absolute bias (log-transformed) shows that absolute bias peaks at 50\% cropland cover for all but the GlobCover map (which continued to increase with cropland cover), and is lowest when the landscape is dominated either by cropland or other cover types (Fig. 2). In other words, bias is highest when cropland cover is mixed evenly with other cover types.

\vspace{-1 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.4\textwidth]{figures/biases_md_lnorm_gam_mu0.pdf}}
\caption{The relationship between the mean absolute bias in cropland maps and cropland cover within agricultural areas (reference map pixels having $>$0.5\% cropland), averaged within the boundaries of magisterial districts (n = 345), as fit with a generalized additive model. Prediction curves are color-coded to the different cropland maps, with the solid line indicating  predicted absolute bias, and the lighter shading the standard error of the coefficients. Models were also fit to the mean absolute biases across all four maps (black curve), and from all maps exlcuding GlobCover (grey surve).}
\label{afoto}
\end{figure}

\subsection{The impact of bias on calculating carbon stocks}
We used the data of \cite{ruesch_new_2008} to calculate carbon densities for African forests, shrublands, croplands, grasslands, and sparse habitats. We assigned the cropland carbon values to map cells in proportion to their cropland cover. For the non-cropland proportions, we assigned the carbon value from each of the other types, such that we created four  different carbon maps for each landcover map at each aggregation scale (Fig. S4), which allowed us to test how carbon estimates vary as a function of i) bias in the cropland estimates and ii) the characteristics of adjacent cover types.

At the 1 km scale, the differences between country-wide total carbon stocks made using any of the cropland maps and the reference were within +/-3\%, regardless of which cover type was adjacent to cropland (Table S1), because differences between total stock estimates are diluted by the large area ($\sim$50-70\%) of the country having no cropland. Comparing the carbon stocks from each map's agricultural area reveals much greater differences (Table S1). SA-LC overestimates carbon stocks by just 2\% when the adjacent cover type is forest, and up to 15\% when it is spare cover. MODIS ranges from negligible overestimates in denser carbon classes (forest, secondary forest, and shrublands) to 8-13\% underestimates for the grassland and sparse classes. GeoWiki underestimates for all types, from $<$1\% for sparse cover to 8\% for forest. GlobCover shows the largest total carbon stock bias, with over-estimates ranging from for sparse lands to 162\% for forest. The magnitude of this bias is due to false positives--GlobCover identified cropland in nearly 50\% of pixels, compared to 30\% for the other three cropland maps.

The spatial patterns of carbon bias (Fig. S4) are a reflection of the cropland biases (Fig. 1). Where cropland is underestimated and the surrounding cover type is of higher carbon density than cropland, carbon density is overestimated. For lower density (grassland and sparse vegetation), there may be slight underestimates.  These tendencies are reflected in the average of each map's spatial biases over agricultural areas (Fig. 3).  For example, MODIS and GlobCover overestimate carbon by $\sim$50\% on average at 1 km resolution when forest is the cropland-adjacent cover (stars in semi-transparent green and gold bars, Fig. 3; Table S2). When it is sparse vegetation (open circles in Fig. 3), MODIS underestimates by nearly 3\% at all scales, whereas maps that overestimate cropland (e.g. SA-LC, semi-transparent red) overestimate carbon density for this cover.  Overall, GeoWiki has the lowest average spatial biases, for all cover types and all resolutions. It's maximum bias is a 12\% overestimate at 1 km for forest cover, thereafter all biases are less than 10\%, and converge to +/-2\% above 1 km aggregation (Fig. 3, Table S2). All maps' biases are within +/-10\% bias after aggregation to 25 km.

\vspace{-0.9 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.4\textwidth]{figures/carbon_veg_scale3.pdf}}
\caption{The average spatial biases in carbon density estimates based on cropland maps, expressed as a percent difference relative to the reference map. The values of actual bias (incorporating bias direction and magnitude) and absolute bias (magnitude only) are presented in semi-transparent and solid colors, respectively, with colors denoting the specific cropland map and symbols indicating which cover type was used to calculate cropland adjacent carbon density. The bar represents the mean biases calculated across each of the 5 cover types. Shrubland and grassland bias values are near zero and not shown for display clarity.}
\label{afoto}
\end{figure}

The patterns of mean absolute spatial bias (solid colored bars in Fig. 3) generally follow that of actual bias, but with somewhat higher magnitudes and a few notable differences. The most notable is that Geo-Wiki has fairly large absolute bias at 1 km, averaging 27\% across across cover types (line in solid blue bar, Fig. 3, Table S2), which is close to the 36-37\% for GlobCover and MODIS. SA-LC has the lowest absolute bias across scales, averaging (across cover types) from 14\% at 1 km to 3\% at 100 km. (Fig. 3, Table S2). The increase in Geo-Wiki's absolute spatial bias relative to SA-LC's can be attributed to the spatially heterogeneous nature of its cropland biases (Fig. 1) compared to the other three cropland maps.

\subsection{Bias in harvested areas, yield, and production estimates}

\begin{itemize}
  \item Used provincial boundaries to extract total cropland from reference data. Cropland totals used to adjust cropland fractions falling within those boundaries from each cropland map so cropland map totals equal the provincial-level total from the reference (SI). 
  \item Disaggregated magisterial district level maize yield and maize harvested area onto the adjusted cropland fractions, and onto the reference map.
  \item Calculated differences in spatial biases of yields and total production estimates at different scales of aggregation. 
  \item Yields aggregated to higher scale using two methods: 1) weighted mean aggregation, where harvested area fractions used to weight yields at each aggregation scale, technically more difficult; 2) mean of any harvested area contributing to aggregation, without weighting. Production aggregated by either calculating production at 1 km and then aggregating to each scale and differencing, or aggregating harvested fractions and type 2 yield, and calculating production at each level.
  \item Average bias normalized as a percentage of mean pixel-level yield or production calculated from the reference map.   
 \end{itemize}

\vspace{-1 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.4\textwidth]{figures/yield_bias_scale.pdf}}
\caption{The mean spatial biases in disaggregated maize yield estimates, and the bias in total crop production estimates calculated from these yields and constrained crop area estimates. The values of actual bias (incorporating bias direction and magnitude) and absolute bias (magnitude only) are presented in semi-transparent and solid colors, respectively, with colors denoting the specific cropland map and symbols indicating whether the bias is related to one of two yield aggregation methods and associated crop production estimates.}
\label{afoto}
\end{figure}

\begin{itemize}
 \item GlobCover average bias shows large yield overestimates (nearly 60\%) at 1 km, declining with aggregation scales. Interestingly, MODIS and GeoWiki start with small yield underestimates that grow to 30\% yield under-estimates at 10 km, declining thereafter. SA-LC has lowest, +/-5\%. (Fig. 4) 
 \item No production biases.
 \item Absolute biases in both yield and production are large. All $>$40\% for both production and yield at 1 km, except SA-LC which is between 15-20\%.  The biases decline with aggregation, but remain above 10\% for yield for most datasets at 100 km.

\end{itemize}

\subsection{Bias in evapotranspiration estimates}
\begin{itemize}
 \item Spatial biases most pronounced for VIC run with MODIS and GlobCover. These two overestimate ET along the southern margin, but typically by less than 5-10\%. (Fig. 5).  In the interior, VIC-GlobCover switches between under-  and over-estimation and then back again moving from West to East, but the bias is on the order of 5\% magnitude.  MODIS shows a similar pattern.    
 \item GeoWiki and SA-LC shows much less discrepancy in interior. Greatest tendency is to underestimate along the southern coast. 
 \item Difference in country-level ET estimates negligible ($<$2\%).
 \item \textcolor{red}{\textbf{Justin}, can you give some thoughts on reasons for these patterns? The tendency seems to be that where cropland is underestimated (blue areas in Fig. 1) then ET is overestimated (red areas in Fig. 5). Here I calculate all difference as (ref - lc) / ref * 100, with lc being one of the four landcover maps. My expectation was that crop ET was higher than most surrounding vegetation types, so one would expect ET to be underestimated where cropland is underestimated?}
\end{itemize}

\vspace{-0.5 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.5\textwidth]{figures/et_bias_map.pdf}}
\caption{Differences in evapotranpiration estimates from a 29 year run of the VIC land surface hydrology model when initialized with LAI response curves derived from the reference map, versus those from the four cropland maps. Columns indicate differences between ET estimates from different portions of the model's time series: annual mean ET; highest annual ET value in the time series (TS max); smallest annual ET value in the time series (TS min); total of mean monthly ET for the three months having the highest ET during the year (Peak 3); the mean monthly ET of the peak ET month.}\label{afoto}
\end{figure}

@@

\begin{itemize}
  \item We selected four magisterial districts with similar climates, and differing abundances of cropland ranging from X-Y\%, and differing patterns of map bias (Fig. S8). 
  \item Using household survey statistics from a project in Zambia that provides information on the average field area of farming households (X ha, SD = Y ha), we assigned each district a number of households determined by the total area of reference cropland in each district divided by mean farm size. \textcolor{red}{\textbf{Tom, Peng}, please correct as needed--maybe there is some variability to account for agent heterogeneity drawn by resampling from the distribution?}. This number of households represents the ``true'' number of households in the district.
  \item Disaggregated cropland to 100 m resolution, which is required by the model. 
  \item An agent-based model was then initialized wherein the household locations were seeded throughout the district, and assigned the required amount of cropland with a radius of X from the household location.
  \item These steps were repeated for each cropland map using the same number of household agents. 
  \item We then examined the differences that occurred between the agent initialization from the reference map and each of the four cropland maps, in terms of the percent of households that did not have cropland allocated, and the amount of cropland left unallocated in each district (Fig. 6).  
  \item The percent households without farms is 0 when cropland is overestimated (because there is plenty of land to choose from), but increases in a 1:1 relationship with the amount of cropland underestimation (positive values on X-axis in Fig. 6). \end{itemize}

\vspace{-0.5 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.3\textwidth]{figures/agent-bias.pdf}}
\caption{Biases in agent-based model initialization relative to cropland map biases, measured in terms of the percent of households having no cropland allocated (top) and the percent of cropland left unallocated (bottom). Dot sizes represent a district number (smallest = district 1; largest = district 4), colors represent the landcover map.  \textcolor{red}{to fix: add x-axis legend, add district number to legend.}}
\label{afoto}
\end{figure}

\begin{itemize}
  \item The percent of land unallocated, however, shows a U-shaped relationship with cropland biases, declining to near zero when cropland bias is between zero and up to $\sim$50\% under-estimated, thereafter increasing. Thus in cases where cropland bias insufficient farmland for households, a signficant share of available cropland can go unallocated (nearly 15\%).  \textcolor{red}{\textbf{Tom, Peng}, I assume this has to do something with seeding and the spatial pattern of cropland, but not sure exactly what. Model not sure which HH to give the land too because there are too many for the amount of cropland available, so it gives up? }
\end{itemize}

@@

\section{Discussion}

Some points to make thrown out now because out of time. Others will be added. Please add any important implications/considerations you see from the results.

Some points from my notes last year: 
\begin{itemize}
  \item Bias as a function of scale
    \begin{enumerate}
      \item At 1 km resolution all landcover products are still fairly biased.  
       \item Bias drops to acceptable levels quickly for geowiki--at 5X5 km, mean bias is just 1\% (overestimated). The absolute bias for this dataset is 10\% or lower from 10X10 km resolution and coarser.  
       \item The SA dataset's bias is fairly consistent but low across all levels of aggregation, amounting to no more than an 8\% overestimate of cropland with absolute bias of similar magnitude. 
       \item MODIS and GlobCover biases (mostly of underestimation) do not dissipate until the higher levels of aggregation. MODIS's actual bias (under-estimation) falls below 10\% at 20 km resolution, but the absolute bias remains above 10\% until more than a 100-fold aggregation is done ($>$100 km resolution).  For GlobCover, it is still too high.  
    \end{enumerate}
  \item Bias as a function of cropland cover
    \begin{enumerate}
      \item Classification algorithms are thus more error-prone where landcover is mixed/heterogenous.
      \item The exception to this lies in the GlobCover dataset, where bias primarily increases as a function of cropland cover. The reason for this is that GlobCover's cropland classes do not provide for 100\% cropland cover, so aggregation tends to exacerbate underestimates. 
      \item Thus caution is needed when aggregating a mixed pixel class.  
        \begin{itemize} 
          \item An example illustrats this: take 4 1 km pixels, 2 of which are 100\% cropland, 2 of which are other cover types. Imagine a landcover product classifies 3 of these as cropland (2 correct, 1 an error of commission), using a cropland class that is defined as 50\% cropland. Aggregating the actual fraction by a factor of 4 will result in a new 4 km pixel having 50\% cropland, whereas aggregating the landcover product's pixels will give just 38\% cropland, even when factoring in the incorrect classification.)
          \end{itemize}
    \end{enumerate}
  \item Bias as a function of method
      \begin{enumerate}
        \item Higher resolution and ensemble-based approaches have less bias.
        \item geowiki represents a fusion of multiple coarse resolution data sources that has undergone extensive validation using a crowdsourcing approach
        \item the SALC dataset is based on 30 m landsat data, but incorporates a range of ancillary data and expert judgement
        \item MODIS and GlobCover data are effectively single source/single algorithm.
        \item \textbf{Newer points begin here}
        \item Statistically constrained constrained landcover estimation approaches provide accurate area-based inferences when aggregated. But spatial errors are still high, as seen with GeoWiki and production/yield estimates. Using these to identify yield gaps at specific map locations is inappropriate, or even for a larger location if it does not coincide with the geographic boundaries of the statistical unit.  
        \item Constrained estimates are also dependent on the accuracy of the statistics.  
    \end{enumerate}
  \item Fix above to have section on bias for global change studies
    \begin{enumerate}
      \item Scales at which it is safe to estimate values of say carbon stocks. 
      \item Above point about bias in disaggregated yield estimates - no point mapping these out. A new approach might be to take these statistically reported yields and then combine them with satellite data to estimate yield variability within the district.  That way would have meaningful reason for disaggregating yields, and would be pegged to real yield values, which would help minimize errors in remote sensing of yields. 
      \item Something on ET - doesn't seem to matter much, but land-atmosphere interactions can make these discrepancies meaningful, particularly since biases occur in arid areas where a lot of irrigation happens--can cause significant impacts on regional climate.  etc. etc. Also we didn't change out land cover types, and the vegetation in SA around the cropland will have reasonably similar LAI and ET responses (I think), thus impact more muted than it might be elsewhere (e.g. in forested landscapes).  
      \item Agent-based models. \textcolor{red}{Tom, Peng, something of significance/implications of this, please}
    \end{enumerate}
  \item Will need a section on way forward for data, etc. Key role of accurate landcover, particularly agricultural.  New methods, vectorized field boundaries seem to be highly valuable, Mapping Africa, Stephanie's paper, Geo-Wiki, etc are the way ahead.

\end{itemize}

Mixed landscapes increase the chances of omission and commission errors by increasing the number of cover classes, or because such landscapes are less spectrally distinct \cite{estes_diylandcover:_2015}
% $\theta$ changes from 0 to 1. (see Figure \ref{afoto}).That means we are considering $\theta$  of the form
%For these solutions we have the following

@@

Describe weighted mean bias reasons.
